---
title: R Tutorial (4)
categories:
  - study
  - deveopment
tags:
  - R
output: 
  md_document:
    preserve_yaml: true
    toc: false
    fig_retina: 2
---

In this post, we will continue our journey down the R road. Specificaly, today's stop is going to be data frames. Visualizations were fun, but to create informative visualizations, good data must come first. Fortunately, R is great for data analysis and wranging, especially thanks to the `dplyr` package, which is R's equivalent of Python's `pandas`. 

# Setup

Let's begin by loading `dplyr`. 

```{r}
library(tidyverse)
```

We will also be using the `nycflights13` package, which is a dataset documenting all flights departing New York City in 2013. 

```{r}
library(nycflights13)
```

Lets take a look at what this dataset looks like.

```{r}
head(flights)
```

# Basic Operations

In this section, we will take a look at some basic operations we can perform on data frames, namely `filter`, `arrange`, `select`,  `mutate`, and `transmute`. If you are familar with SQL or `pandas`, the semantics of some of this words might come a bit more naturally. But even if they don't, worry not; we will go through each function one by one and get our hands dirty.

## Filter



















